---
title: "CLASSIFICATION OF ALZHEIMER’S DISEASE USING MACHINE LEARNING APPROACH"
name: "Anthony Omowumi"
BNumber: "B00892305" 
CourseWork: "Machine Learning and Data Modelling - COM737" 
output:
  html_document: default
  word_document: default
date: "2023-04-26"
---

```{r}
options(knitr.duplicate.label = "allow")
knitr::opts_chunk$set(echo = TRUE)
```

#1.0 IMPORTING NECESSARY LIBRARIES FOR THE ANALYSIS
```{r}

library(editrules)
library(skimr)
library(randomForest)
library(dplyr)
library(data.table)
library(Publish)
library(MLmetrics)
library(smotefamily)
library(ROCR)
library(caTools)
library(ggplot2)
library(corrplot)
library(Boruta)
library(missForest)
library(caret)
library(e1071)

```

#1.1 IMPORTING YOUR DATASET INTO R
```{r}
AIBL <- read.csv("AIBL.csv",header=T, stringsAsFactors = F)
AIBL
```
#1.2 Performing Descritpive Statistics about the dataset
```{r}
#View(AIBL) #To open and view the dataset in a dataframe format opening as a table
names(AIBL) #To check for column names
str(AIBL) #To check the structure of the dataset
sapply(AIBL, class) #Viewing the column names and datatypes
dim(AIBL) #This shows the dimension of the dataset in columns and rows
summary(AIBL) #Gives the statistical summary of the dataset in terms of min,max, median, 1Q and 3Q.
```

#1.3 Conducting data pre-processing on the dataset
```{r}
AIBL <- subset(AIBL, select = -c(RID)) #This column was dropped because it is just a numbering column not relevant to the study
AIBL_1 <- AIBL #Saving it into another data frame in order to perform the necessary data pre-processing and also retent the data provenance

AIBL_1$Diagnosis[AIBL_1$Diagnosis == 3]<- 2 #MCI and AD is merged into one category. 1 for HC (Healthy Control) and 2 for non-HC (Non-Healthy Control) factor.


library(data.table)
setDT(AIBL_1)
AIBL_1[, MMSCORE:= cut(MMSCORE,
            breaks = c(0,9,20,24,30),
            include.lowest = T,
            labels = c("4","3","2","1"))] # 1 belongs to the category normal(30-25), 2 belongs to the category mild/early(24-21), 3 belong to the category moderate(20-10) while 4 mbelongs to the category severe (9-0)

class(AIBL_1$MMSCORE) #Checking the class of the column MMSCORE after the preprocessing
AIBL_1[, table(MMSCORE)] #The variable was grouped into their different categories


library("tidyr")
library("dplyr")
AIBL_1 <- AIBL_1 %>% 
  mutate_if(is.numeric, ~replace(., . == -4, NA)) #Replacing all the columns with -4, missing values with NA

AIBL_1$Diagnosis[AIBL_1$Diagnosis == 7]<- NA #Use to replace 7 which is a missing/inaccurate value with NA

#Storing into another dataframe after the preprocessing has been done.
AIBL1 <- AIBL_1 
AIBL1

sum(is.na(AIBL1)) #To get the sum of na's in the preprocessed data. this is equal to 1258
summary(AIBL1) #MH16SMOK shows the highest number of NA's and also the summary shows that not all the datapoint in the columns belongs to the same scale using the min and max values as reference. Hence, before model fitting the data columns would be scale/normalize.


#Identifying special values in the dataset
is.finite(c(Inf, NaN)) #checking for Special values (Inf, NA and NaN) and correcting it.
is.special <- function(x){
  if (is.numeric(x)) !is.finite(x) else is.na(x)
} #special characters. function define

sapply(AIBL1, is.special)  #Special character detection in the dataset.
AIBL1[mapply(is.special, AIBL1)] <- NA #Applying NA to any special characters if present.


#Showing the outliers using boxplot.
Boxplot_Outliers_Detection <- subset(AIBL1, select = -c(Diagnosis))
boxplot(Boxplot_Outliers_Detection,cex.axis=.3, col="gray",horiz=F, dotplot=FALSE) #The boxplot shows that BAT126 has the highest outliers while others are mild, still within the Interquartile range.


```

#2.0 Performing a Missing data imputation using RandomForest Algorithm
```{r}
AIBL1 <- AIBL1 %>% mutate(across(c(PTGENDER, MHPSYCH, MH2NEURL, MH4CARD, MH6HEPAT, MH8MUSCL, MH9ENDO, MH10GAST,MH12RENA, MH16SMOK, MH17MALI, APGEN1, APGEN2, CDGLOBAL,MMSCORE,LIMMTOTAL, LDELTOTAL, Diagnosis ), as.factor)) #Defining the data type of the feature column

str(AIBL1)
dim(AIBL1)

#Perfroming missing values imputation using Random Forest library(missForest)
set.seed(12345) #This allows for reproducibility 
AIBL1.imp <- missForest(AIBL1, verbose = T)
AIBL1.imp$OOBerror #After 3 iteration training, the final NMRSE(True normalized root mean squared error) obtained is 0.4859566 and PFC was 0.2674520. OOBerror is the imputation error estimates made to understand the quality of an imputation.

names(AIBL1.imp)
AIBL1.imp$ximp

summary(AIBL1.imp$ximp)
dim(AIBL1.imp$ximp)
str(AIBL1.imp$ximp)

sum(is.na(AIBL1.imp$ximp)) #zero na value after the imputation

table(AIBL1.imp$ximp$Diagnosis) # Checking for class size after imputation
prop.table(table(AIBL1.imp$ximp$Diagnosis)) #Diagnosis 01is 71% while Diagnosis 2 is 29%
```


#2.1 Visualising the dataset and find for Multicollinearity between the dataset
```{r}
#BOXPLOT for Age of Patient's vs Diagnosis

boxplot(list(AIBL1.imp$ximp$Age[AIBL1.imp$ximp$Diagnosis==1], AIBL1.imp$ximp$Age[AIBL1.imp$ximp$Diagnosis==2]), col = "gray", xlab="Diagnosis", ylab="Patient's Age", main="Boxplot for Patient's Age vs Diagnosis") #The boxplot shows that people with other age bracket are most likely to be in Non-HC category of diagnosis..

hist(AIBL1.imp$ximp$Age, breaks=10, col = "gray", border = "white")
```

#2.2 Correlation checks within the Features before Variable Feature selection
```{r}
AIBL2 <- AIBL1.imp$ximp #Saving AIBL.imp as AIBL2

table(sapply(AIBL2, is.factor)) #This shows the number of factor columns in the dataframe
AIBL2 <- sapply(AIBL2, as.numeric)
sapply(AIBL2, class)

AIBL2 <- as.data.frame(AIBL2)

#Correlation plot only works with numerical value that is why all the non-numerical value is coerced into a numerical values
library(corrplot)
CorrMatrix <- round(cor(AIBL2[,1:31]),2)
CorrMatrix

corrplot(CorrMatrix, method = 'circle', order = "hclust", addrect = 4)

CorrMatrix[CorrMatrix < 0.8 | CorrMatrix ==1] <- "" #This shows there is high correlation between CDGLOBAL, LIMMTOTAL, LDELTOTAL, DIAGNOSIS, HMT40, HMT3. Further analysis would be conducted to find the important features for the analysis(Variable selection)
```

#2.3 Feature Selection using Boruta Algorithm 
```{r}
library(Boruta)
library(mlbench)
library(caret)
library(randomForest)

set.seed(123)
AIBL2_Boruta <- Boruta(Diagnosis ~., data = AIBL2, doTrace=2, maxRuns = 500) #9 attributes were choosen as important

print(AIBL2_Boruta)
plot(AIBL2_Boruta, las=2, cex.axis=0.5) #Computes the z scores of the Confirmed, Rejected and Tentative attributes in the selection

plotImpHistory(AIBL2_Boruta)

#FIXING THE TENTATIVE ATTRIBUTE
#The tentative attributes will be classified as confirmed or rejected by comparing the median Z score of the attributes with the median Z score of the best shadow attribute.

final_bor <- TentativeRoughFix(AIBL2_Boruta)
print(final_bor)
plot(final_bor, las=2, cex.axis =.5)#Shows that 12 attributes has been choosen as important features.

getSelectedAttributes(final_bor, withTentative = F)
attStats(AIBL2_Boruta)


SelectedAttributes <- c(1, 4, 13, 17, 20, 22, 25, 26, 27, 28, 29, 30, 31) #Selected Boruta algorithm for important feature and diagnosis(response variable/predictor)

AIBL2_new <- AIBL2[, SelectedAttributes] %>% mutate(across(c(PTGENDER, MH2NEURL, APGEN1, CDGLOBAL,MMSCORE,LIMMTOTAL, LDELTOTAL, Diagnosis), as.factor)) #Coercing the data type back for model building and adding the diagnosis column to it.

class(AIBL2_new)
names(AIBL2_new)
dim(AIBL2_new)
str(AIBL2_new)


#Scale the 5 numeric features in the selected attribute data frame
AIBL2_new$HMT3 <- (AIBL2_new$HMT3 - min(AIBL2_new$HMT3)) / (max(AIBL2_new$HMT3) - min(AIBL2_new$HMT3))
AIBL2_new$HMT40 <- (AIBL2_new$HMT40 - min(AIBL2_new$HMT40)) / (max(AIBL2_new$HMT40) - min(AIBL2_new$HMT40))
AIBL2_new$HMT102 <- (AIBL2_new$HMT102 - min(AIBL2_new$HMT102)) / (max(AIBL2_new$HMT102) - min(AIBL2_new$HMT102))
AIBL2_new$RCT20 <- (AIBL2_new$RCT20 - min(AIBL2_new$RCT20)) / (max(AIBL2_new$RCT20) - min(AIBL2_new$RCT20))
AIBL2_new$RCT392 <- (AIBL2_new$RCT392 - min(AIBL2_new$RCT392)) / (max(AIBL2_new$RCT392) - min(AIBL2_new$RCT392))

str(AIBL2_new)
#View(AIBL2_new)

#Boruta performed 499 iterations in 2.031299 mins.
#9 attributes confirmed important: APGEN1, CDGLOBAL, HMT102, HMT3, HMT40 and 4 more;
#18 attributes confirmed unimportant: Age, APGEN2, AXT117, BAT126, HMT100 and 13 more;
#3 tentative attributes left: MH2NEURL, PTGENDER, RCT392;
```

#2.4 Checking data Proportion and balancing the data using SMOTE
```{r}
table(AIBL2_new$Diagnosis) # Checking for class size 1 is 609(HC) and Non-HC is 253
prop.table(table(AIBL2_new$Diagnosis)) #Outcome 1 is 71% while outcome 2 is 29%. This is an imbalance data.

AIBL2_new <- sapply(AIBL2_new, as.numeric)
sapply(AIBL2_new, class)

AIBL2_new <- as.data.frame(AIBL2_new)

#SMOTE was used to balance the Outcome class
Diagnosis_balancing <- SMOTE(AIBL2_new[-13], 
              as.numeric(AIBL2_new$Diagnosis),
              K = 3, dup_size = 1) 
str(Diagnosis_balancing)

AIBL.new.data <- bind_cols(Diagnosis_balancing$data[13], Diagnosis_balancing$data[-13])
str(AIBL.new.data)

# Make dependent variable as a factor (categorical).
AIBL.new.data$class = as.factor(AIBL.new.data$class)
#Set the label colname back to "Outcome".
names(AIBL.new.data)[1]<- "Diagnosis"
names(AIBL.new.data)

# Check types of variables and class sizes.
str(AIBL.new.data)
class(AIBL.new.data$Diagnosis)
table(AIBL.new.data$Diagnosis)
prop.table(table(AIBL.new.data$Diagnosis)) #After Smote has been done class 1 is now 55% and class 2 is now 45%

barplot(table(AIBL.new.data$Diagnosis), main = "Diagnosis Count After SMOTE", xlab = "Diagnosis", ylab= "Count", border = "white", col="gray")

```

#3.0 SPLITTING DATASET INTO TRAINING AND TESING

```{r}

# Splitting the dataset into the Training set and Test set
# install.packages('caTools')
AIBL.new.data$Diagnosis <- as.factor(AIBL.new.data$Diagnosis)
set.seed(123)
split = sample.split(AIBL.new.data$Diagnosis, SplitRatio = 0.7)
AIBL_training = subset(AIBL.new.data, split == TRUE)
AIBL_test = subset(AIBL.new.data, split == FALSE)
dim(AIBL_training) #780 13
dim(AIBL_test) #335 13
AIBL_training$Diagnosis<- as.factor(AIBL_training$Diagnosis)
AIBL_test$Diagnosis<- as.factor(AIBL_test$Diagnosis)

```

#3.1 MODEL FITTING USING RANDOM FOREST ALGORITHM - TRAINING & TESTING
```{r}
#Random Forest

# Train and tune the random forest (rf) algorithm on the training data.
library(randomForest)

# Find the optimal value of mtry.
set.seed(123) 
mtry <- tuneRF(AIBL_training[-1],AIBL_training$Diagnosis, ntreeTry=500,
               stepFactor=1.5,improve=0.01, trace=TRUE, plot=TRUE)

best.m <- mtry[mtry[, 2] == min(mtry[, 2]), 1] #Is used to compute the best mtry for RF algorthm

print(mtry) 
print(best.m)

#Apply random forest (rf) with the optimal value of mtry.
set.seed(123)
random_forest_model<-randomForest(Diagnosis~.,data=AIBL_training, mtry=best.m, importance=TRUE,ntree=500) #Training the RF with the optimal value of mtry..
print(random_forest_model) #4.36% error rate.  it means there was 4.49% out of Bag error rate..
plot(random_forest_model)

importance(random_forest_model)
varImpPlot(random_forest_model)

library("caret")
library(MLmetrics)

pred <- predict(random_forest_model, newdata = AIBL_test) #Predicting the Test set results.

cm <- ConfusionMatrix(pred, AIBL_test$Diagnosis) # Storing the Confusion Matrix
(Classification.Accuracy <- 100*Accuracy(pred, AIBL_test$Diagnosis))# Model Accuracy = 97.01%
l <- table(AIBL_test$Diagnosis, pred) 
confusionMatrix(l, mode = "everything")# Recall = 96.26% and F1 Score = 97.30%

#Prediction and Calculate Performance Metrics
pred_1<-predict(random_forest_model,newdata = AIBL_test,type = "prob")

#library(ROCR)
(p <- prediction(pred_1[,2], AIBL_test$Diagnosis))

# 0. Accuracy.
(acc = performance(p, "acc"))
plot(acc,main="Accurcay Curve for Random Forest",col=2,lwd=2)

# 1. Area under curve
auc <- performance(p, "auc") #The AUC is 98.99%
auc@y.values[[1]]

# 2. True Positive and Negative Rate
pred_3 <- performance(p, "tpr","fpr")

# 3. Plot the ROC curve
plot(pred_3,main="ROC Curve for Random Forest",col=2,lwd=2)
abline(a=0,b=1,lwd=2,lty=2,col="purple")

```

#3.2 MODEL FITTING USING SUPPORT VECTOR MACHINE
```{r}
#Using SVM
#1. FITTING THE MODEL WITHOUT setting the gamma and cost(regularization)
# Fit the SVM model with the radial kernel using gamma and cost equals to 1. 
library(e1071)

set.seed(123)
svm.model.radial <- svm(Diagnosis ~ ., data = AIBL_training, kernel = 'radial', gamma=1, cost=1)
svm.model.radial
summary(svm.model.radial)

# Find false positives and false negatives for the radial SVM
table(Prediction = predict(svm.model.radial, AIBL_training),Truth = AIBL_training$Diagnosis)
#Predicting using test dataset
y_pred = predict(svm.model.radial, newdata = AIBL_test)
y_pred
#Confusion Matrix
cm1= table(AIBL_test[, 1], y_pred)
cm1 
confusionMatrix(cm1, mode = "everything")


#Tuning the regularization and Gamma but using the same Kernel RBF
set.seed(123)
tuned.svm <- tune(svm, Diagnosis ~., data = AIBL_training,kernel="radial",ranges=list(cost=c(0.001, 0.01, 0.1, 1,5,10,100)), gamma=c(0.5,1,2,3,4))
tuned.svm
summary(tuned.svm)

# Showing the best model obtained
bestmodel <- tuned.svm$best.model
summary(bestmodel)

y_pred.tuned= predict(bestmodel, newdata = AIBL_test)
y_pred.tuned
cm= table(AIBL_test[, 1], y_pred.tuned)
cm

confusionMatrix(cm, mode = "everything")
```

#3.3 MODEL FITTING USING K-NEAREST NEIHGBOUR
```{r}
#Using KNN
library(mlbench)

# Fit the model on the training set
set.seed(123)
knn.model <- train(
  Diagnosis ~., data = AIBL_training, method = "knn",
  trControl = trainControl("cv", number = 10),
  tuneLength = 20
)

# Plot model accuracy vs different values of k
plot(knn.model)

# Print the best tuning parameter k that
# maximizes model accuracy
knn.model$bestTune

# Make predictions on the test data
predicted.classes <- knn.model %>% predict(AIBL_test)
head(predicted.classes)

m <- table(AIBL_test$Diagnosis, predicted.classes) 
confusionMatrix(m, mode = "everything") 

# Compute model accuracy rate
mean(predicted.classes == AIBL_test$Diagnosis)
```

#3.4 UNSUPERVISED LEARNING USING K-MEANS
```{r}
#Doing Kmeans Clustering
#install.packages("stats")
#install.packages("ggfortify")
library("stats")
library("ggfortify")


mydata = select(AIBL.new.data, c(2:13)) #unlabelling was done for dataset because we are dealing with unsupervised learning. 
View(mydata)

#wss plot function
wssplot <- function(data, nc=15, seed=1234)
{
  wss<- (nrow(data) -1) * sum(apply(data, 2, var))
  for (i in 2:nc) {
    set.seed(seed)
    wss [i] <- sum(kmeans(data, centers = i)$withinss)
  }
  plot(1:nc, wss,type="b", xlab="Number of clusters", ylab="Within groups sum of squares")
  
}

wssplot(mydata)#To chosen the optimal values

#k-Means Cluster
KM.1 <- kmeans(mydata, 2) #The elbow method gives a optimal value between 2 and 3. So 2 was chosen 
KM.1
#Cluster Plot
autoplot(KM.1, mydata, frame=TRUE) 

KM.2 <- kmeans(mydata, 3) 
#Cluster Plot
autoplot(KM.2, mydata, frame=TRUE)


#Evaluating Cluster Analysis
KM.1$centers #The centers of all the features indicates they are distinct in nature and no overlap. Looking at the plot, it shows the same distinctiveness and no overlapping of the faetures. 

table(KM.1$cluster, AIBL.new.data$Diagnosis)

```



